Bayesian Deep Gaussian Processes for Correlated Functional Data: \\ A Case Study in Cosmological Power Spectra

### 2025-04-22 paper/main.tex
minor edits in abstract

\title{Bayesian Deep Gaussian Processes for Correlated Functional Data: \\
        A Case Study in Cosmological Power Spectra}
\author{Stephen A. Walsh\thanks{Corresponding author: Division of Natural Sciences, 
        Math and Technology, Elms College, {\tt walshst@elms.edu}} \and 
        Annie S. Booth\thanks{Department of Statistics, North Carolina State University} \and
        David Higdon\thanks{Department of Statistics, Virginia Tech} \and
        Marco A.R. Ferreira\footnotemark[3] \and
        Jared Clark\footnotemark[3] \and
        Kelly Moran\thanks{Los Alamos National Laboratory} \and
        Katrin Heitmann\thanks{Argonne National Laboratory}}
\date{\today}

@@

\begin{abstract} 
Understanding the structure of our universe and the distribution of matter is an 
area of active research.  As cosmological surveys grow in complexity, the development 
of emulators to efficiently and effectively predict matter power spectra is essential.  
We are particularly motivated by the Mira-Titan Universe simulation
suite which, for a specified cosmological parameterization (termed a ``cosmology''), 
provides multiple response curves of various fidelities, including correlated 
functional realizations.  Our objective is two-fold.  First, we estimate 
the underlying true matter power spectra, with appropriate uncertainty 
quantification (UQ), from all of the provided curves.  To this end, we propose a 
novel Bayesian deep Gaussian process (DGP) hierarchical model which synthesize 
all the information and estimates the underlying matter power spectra, while providing 
effective UQ.  Our model extends previous work on Bayesian DGPs from scalar responses 
to functional ones.  Second, we leverage our predicted power spectra from various 
cosmologies in order to accurately predict the entire matter power spectra for an 
unobserved cosmology.  For this task, we leverage basis function representations 
of the functional spectra to train a separate Gaussian process emulator.  
Our method performs well in synthetic exercises and against the benchmark cosmological 
emulator (Cosmic Emu).  We provide reproducible code in a public git repository.
\end{abstract}

>>>>

\title{Bayesian Deep Gaussian Processes for Correlated Functional Data: \\
        A Case Study in Cosmological Power Spectra}
\author{Stephen A. Walsh\thanks{Corresponding author: Division of Natural Sciences, 
        Math and Technology, Elms College, {\tt walshst@elms.edu}} \and 
        Annie S. Booth\thanks{Department of Statistics, Virginia Tech} \and
        David Higdon\footnotemark[2] \and
        Marco A.R. Ferreira\footnotemark[2] \and
        Jared Clark\footnotemark[2] \and
        Kelly Moran\thanks{Los Alamos National Laboratory} \and
        Katrin Heitmann\thanks{Argonne National Laboratory}}
\date{\today}

@@

\begin{abstract} 
Understanding the structure of our universe and the distribution of matter is an 
area of active research.  As cosmological surveys grow in complexity, the development 
of emulators to efficiently and effectively predict matter power spectra is essential.  
We are particularly motivated by the Mira-Titan Universe simulation
suite which, for a specified cosmological parameterization (termed a ``cosmology''), 
provides multiple response curves of various fidelities, including correlated 
functional realizations.  Our objective is two-fold.  First, we estimate 
the underlying true matter power spectra, with appropriate uncertainty 
quantification (UQ), from all of the provided curves.  To this end, we propose a 
novel Bayesian deep Gaussian process (DGP) hierarchical model which synthesizes 
all the simulation information to estimate the underlying matter power spectra
while providing effective UQ.  Our model extends previous work on Bayesian DGPs 
from scalar responses to functional ones.  Second, we leverage our predicted 
power spectra from various cosmologies in order to accurately predict the entire 
matter power spectra for an 
unobserved cosmology.  For this task, we use basis function representations 
of the functional spectra to train a separate Gaussian process emulator.  
Our method performs well in synthetic exercises and against the benchmark cosmological 
emulator (Cosmic Emu).
\end{abstract}
